High-Throughput Computing on \\ High-Performance Platforms: A Case Study

### 2017-10-28 arxiv/v2/abstract.tex
Update arxiv version

The computing systems used by LHC experiments has historically consisted of
the federation of hundreds to thousands of distributed resources, ranging
from small to mid-size resource.  In spite of the impressive scale of the
existing distributed computing solutions, the federation of small to mid-size
resources will be insufficient to meet projected future demands. This paper
is a case study of how the ATLAS experiment has embraced Titan -- a DOE
leadership facility in conjunction with traditional distributed high-
throughput computing to reach sustained production scales of approximately
52M core-hours a years. The three main contributions of this paper are:  (i)
a critical evaluation of design and operational considerations  to support
the sustained, scalable and production usage of Titan;  (ii) a preliminary
characterization of a next generation executor for PanDA to support new
workloads and  advanced execution modes; and (iii) early lessons for how
current and future experimental and observational systems can be integrated
with production supercomputers and other platforms in a general and
extensible manner.

>>>>

The computing systems used by LHC experiments has historically consisted of
the federation of hundreds to thousands of distributed resources, ranging
from small to mid-size resource.  In spite of the impressive scale of the
existing distributed computing solutions, the federation of small to mid-size
resources will be insufficient to meet projected future demands. This paper
is a case study of how the ATLAS experiment has embraced Titan---a DOE
leadership facility in conjunction with traditional distributed high-
throughput computing to reach sustained production scales of approximately
52M core-hours a years. The three main contributions of this paper are:  (i)
a critical evaluation of design and operational considerations  to support
the sustained, scalable and production usage of Titan;  (ii) a preliminary
characterization of a next generation executor for PanDA to support new
workloads and  advanced execution modes; and (iii) early lessons for how
current and future experimental and observational systems can be integrated
with production supercomputers and other platforms in a general and
extensible manner.

### 2017-10-28 arxiv/v2/conclusion.tex
Update arxiv version

% \ifreview
% Comments to address in this section:
% \begin{itemize}
% 	\color{red} 
% 	% \item REVIEWER 2: How does Summit, the TITAN successor, compare to the
% 	% conclusions? The end of the TITAN life-cycle basically invalidates the
% 	% applicability of all conclusions if this generalisation is missing.
% 	% \item REVIEWER 3: This statement in the "conclusions" section is quite
% 	% exaggerated: "The experience captured in this paper will prove invaluable
% 	% for designing WMS for computational campaigns and will provide a baseline
% 	% to evaluate the relative merits of different approaches." WMS systems for
% 	% LHC experiments have been developed and run in production for many years;
% 	% while this paper adds interesting experience and data points, it's
% 	% unlikely that this will be "invaluable" for new designs and a cornerstone
% 	% for their evaluation. \jhanote{I think this comment can also be ignored.}
% \end{itemize}
% \fi

@@

% More generally, this paper highlights the fundamental role of WMS for
% experimental science.

@@

{\bf \footnotesize Acknowledgements: This work is funded by Award number DE-
SC0016280 from the Office of Advanced Scientific Computing Research within the
Department of Energy. This research used resources of the Oak Ridge Leadership
Computing Facility at the Oak Ridge National Laboratory, which is supported by
the Office of Science of the U.S.Department of Energy under Contract No. DE-
AC05-00OR22725. \par}

% The development of NGE is looking forward to deal with these challenges by
% providing an abstraction layer for HPC resources together with a set of
% APIs for the coordination and fast deployment of heterogeneous workloads.
% In this paper, we provided the results of a preliminary testing of the NGE.
% The testing pointed out the robustness of NGE at handling an high number of
% tasks and worker nodes by mantaining an overhead that grows linearly with
% the number of tasks.

% Recent efforts with Google Cloud for the CMS experiment --
% HEPCloud~\cite{hepcloud,googlehep} represent similar yet different
% approaches to dynamic resource federation. As the type, heterogeneity and
% complexity of future platforms  (e.g., Exascale, Commercial Clouds) for
% science increases, there will be a greater need and emphasis on
% abstractions to ensure WMS can both utilize as well as support evolving
% platforms. Our experience demonstrated the advantages arising from an
% abstractions based approach \textendash{} in particular scalable
% implementations of pilot abstraction.

>>>>

@@

@@

{\bf \footnotesize Acknowledgements: This work is funded by Award number
DE-SC0016280 from the Office of Advanced Scientific Computing Research within
the Department of Energy. This research used resources of the Oak Ridge
Leadership Computing Facility at the Oak Ridge National Laboratory, which is
supported by the Office of Science of the U.S. Department of Energy under
Contract No. DE-AC05-00OR22725. \par}

### 2017-10-28 arxiv/v2/introduction.tex
Update arxiv version

% \ifreview
% Comments to address in this section:
% \begin{enumerate}
% 	\color{red} 
% 	\item REVIEWER 1: The introduction is overly verbose and prosaic. The
% 	praise for LHC collaborations is off-putting, especially as it ignores
% 	other work. For example, claiming "CMS and ATLAS experiments utilize
% 	arguably the largest production grade distributed computing solutions"
% 	disregards the obvious commercial contenders. By blatantly excluding
% 	obvious comparisons, it raises the question how much other work is
% 	ignored. \jhanote{Added "academic" after arguably. Any more?}
% \end{enumerate}
% \fi

@@

The LHC workloads are comprised of tasks that are independent of each other,
however, the management of the distribution of workloads across many
heterogeneous resources, the effective utilization of resources and efficient
execution of workloads present non-trivial challenges. Many software
solutions have been developed in response to these challenges. The CMS
experiment, devised a solution based around the
HTCondor~\cite{thain2005distributed} software ecosystem. The
ATLAS~\cite{Aad:2008} experiment utilizes the Production and Distributed
Analysis (PanDA) workload management system~\cite{Maeno2011} (WMS) for
distributed data processing and analysis. The CMS and ATLAS experiments
utilize, arguably the largest academic production grade distributed computing
solutions, and have symbolized the paradigm of {\it high-throughput
computing} (HTC), i.e., the effective execution of many independent tasks.

In spite of the impressive scale of the ATLAS distributed computing system --
in the number of tasks executed, the number of core hours utilized, and the
number of distributed sites utilized,  demand for computing systems will soon
significantly outstrip current and projected supply.   The data volumes that
will need analyzing in LHC-Run 3 ($\approx$2022) and the high-luminosity
era (Run 4) will increase by factors of 10--100 compared to the current phase
(Run 2). There are multiple levels at which this problem needs to be
addressed: the utilization of emerging parallel architectures (e.g.,
platforms); algorithmic and advances in analytical methods (e.g., use of
Machine Learning); and the ability to exploit different platforms (e.g.,
clouds and supercomputers).

% As the LHC prepares for Run 3 in $~\approx$ 2022 and the high-luminosity
% era (Run 4), it is anticipated that the data volumes that will need
% analyzing will increase by factors of 10-100 compared to the current phase
% (Run 2). Data will be larger in volume but will also require more
% sophisticated computational processing.

>>>>

@@

The LHC workloads are comprised of tasks that are independent of each other,
however, the management of the distribution of workloads across many
heterogeneous resources, the effective utilization of resources and efficient
execution of workloads present non-trivial challenges. Many software
solutions have been developed in response to these challenges. The CMS
experiment, devised a solution based around the
HTCondor~\cite{thain2005distributed} software ecosystem. The
ATLAS~\cite{Aad:2008} experiment utilizes the Production and Distributed
Analysis (PanDA) workload management system~\cite{Maeno2011} (WMS) for
distributed data processing and analysis. The CMS and ATLAS experiments
utilize, arguably the largest academic production grade distributed computing
solutions, and have symbolized the paradigm of high-throughput computing
(HTC), i.e., the effective execution of many independent tasks.

In spite of the impressive scale of the ATLAS distributed computing
system---in the number of tasks executed, the number of core hours utilized,
and the number of distributed sites utilized,  demand for computing systems
will soon significantly outstrip current and projected supply.   The data
volumes that will need analyzing in LHC-Run 3 (\(\approx\)2022) and the
high-luminosity era (Run 4) will increase by factors of 10--100 compared to
the current phase (Run 2). There are multiple levels at which this problem
needs to be addressed: the utilization of emerging parallel architectures
(e.g., platforms); algorithmic and advances in analytical methods (e.g., use
of Machine Learning); and the ability to exploit different platforms (e.g.,
clouds and supercomputers).

### 2017-10-28 arxiv/v2/overall.tex
Update arxiv version

\ifreview
Comments to address in this section:
\begin{enumerate}
    \color{red}

    \item REVIEWER 2: What are the reasons for the low efficiency of PanDA
    Broker? Why are only 30\% of backfill slots useable, if they are basically
    uncontested? \jhanote{MT to address this via a few sentences in Section
    IV.}

    % \item REVIEWER 2: Why does Fig 3 show an increase of absolute and
    % relative efficiency in Aug-16, when the text claims this in Sep-16 due
    % to more PanDA Brokers?\mtnote{Fixed.}

    % \item REVIEWER 2: Why are only 20 PanDA Brokers deployed if this is a
    % limiting factor? Why is it questionable that more PanDA Brokers would
    % increase efficiency? If only 5/6 of resources below the mean are
    % unusable, this implies an efficiency of almost 60\% being
    % reachable.\mtnote{From our text: ``further research is ongoing to
    % understand whether an even greater number of brokers would yield even
    % greater efficiency. This research focuses on evaluating the overheads
    % of input/output files staging, including its impact on DTNs, and on an
    % alternative design of PanDA Broker that enables the concurrent
    % submission of multiple MPI scripts''. I believe this is an answer good
    % enough.}

    % \item REVIEWER 2: Given an expected increase in resource need of
    % 10-100, how is the capacity of 4\% WLCG significant? How does this
    % relate to the manpower required to develop and maintain a custom
    % solution per HPC site? How many TITAN resources are actually available
    % to HEP, given the use by other domains? The magnitude of capacities
    % presented as available is in major contrast to the requirements
    % sketched in the introduction.

    % \item REVIEWER 2: How does the poor metadata read performance of
    % 1500-6300s (25\%-100\% of minimum walltime!) affect efficiency? Is it
    % factored into the choice of walltime requests?\mtnote{Thanks to
    % Sergey's input I corrected the figure previously reported.} Is it
    % applicable only to TITAN, to HPC in general, or even to regular WLCG
    % sites?\mtnote{This is clearly out of scope for this paper.} Are there
    % solutions at application, e.g. the use of packaged python scripts and
    % libraries?\mtnote{Added the following: ``While these figures are
    % specific to OLCF and TItan, reengineering on AthenaMP could further
    % improve startup time on every platform.''} The information presented is
    % insufficient to estimate whether this issue applies to other
    % environments and usecases as well, or is simply due to suboptimal
    % implementation/configuration.

    % \item REVIEWER 3: The HEP-SPEC06 benchmark, which is used as a
    % ``computing power'' measure in LCG is mostly integer-artihmetic based:
    % FP operations accounts for \~10\% of the average workload (see [1],
    % section ``Why do you use `all\_cpp' instead of SPECint or SPECfp?'')
    % While this figure certainly comes from averaging over many different
    % workloads across all experiments, it seems a bit surprising that FP
    % scheduling contention is responsible for 50\% decrease in compute
    % efficiency on Titan when compared to other WLCG sites (paper page
    % 6).\mtnote{Thanks to Sergey's input I downplayed the role played by FP
    % and up-played the role played by new CPUs.}
\end{enumerate}
\fi

@@

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \vspace{-0.3in}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and
    CPU core-hours used by ATLAS (blue). GPU core-hours unaccounted for as
    they cannot be used by ATLAS\@. Efficiency of PanDA Brokers defined as
    percentage of total Titan's backfill availability used by ATLAS (Red
    labels).}\label{fig:backfill-utilization}
\end{figure}

@@

% This suggests that the differences in efficiency did not depend on a
% comparable difference in the total backfill availability.

% \jhanote{change last sentence to: ... efficiency does not depend on total
% backfill availability." ??}
% \jhanote{shouldn't we be consistent in the months used to quote efficiency
% and utilization? April 2016 and Feb 2017 vs May 2016 and Feb 2017??}

@@

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/titan_backfill_avail.png}
    \vspace{-0.3in}
    \caption{62555 measures of Backfill availability on Titan during the
    experiment time window. Mean number of work nodes available 691; mean
    walltime available 126 minutes.}\label{fig:backfill-distrib}
\end{figure}

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison_vsquashed.pdf}
    \vspace{-0.3in}
    \caption{Distributions of the time taken to simulate one event when
    placing 2 simulations (h1) or 1 simulation (h2) per Titan's CPU\@. 2
    simulation use 16 cores per node, 1 simulation 8. 50 Events; 1 Titan
    worker nodes; 16 work threads per node; 100 events per
    node.}\label{fig:comparison-8-16cores}
\end{figure}

A performance analysis of Titan's AMD CPUs for detector simulations also helps
to compare Titan and Grid site performance. Usually, Grid sites expose
resources with heterogeneous CPU architectures and 8 (virtual) cores, while
Titan's offer an homogeneous 16 cores architecture. We used the rate of events
processes per minute as a measure of the efficiency of executing the same
detector simulation on Titan or Grid sites. Comparisons of the efficiencies of
Titan to the BNL and SIGNET Grid sites, normalized for 8 cores, show that the
effective performance per-core at Titan is
\(\approx\)0.57 event per minute, roughly 1/2 of BNL and  1/3 of SIGNET.

%performances.

% The CPUs at the Grid sites have one FP scheduler per core and are on
% average newer than the CPU of Titan.

>>>>

@@

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{backfill_consumption.pdf}
    \vspace{-0.3in}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and
    CPU core-hours used by ATLAS (blue). GPU core-hours unaccounted for as
    they cannot be used by ATLAS\@. Efficiency of PanDA Brokers defined as
    percentage of total Titan's backfill availability used by ATLAS (Red
    labels).}\label{fig:backfill-utilization}
\end{figure}

@@

@@

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{titan_backfill_avail.png}
    \vspace{-0.3in}
    \caption{62555 measures of Backfill availability on Titan during the
    experiment time window. Mean number of work nodes available 691; mean
    walltime available 126 minutes.}\label{fig:backfill-distrib}
\end{figure}

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{tx8_tx16_comparison_vsquashed.pdf}
    \vspace{-0.3in}
    \caption{Distributions of the time taken to simulate one event when
    placing 2 simulations (h1) or 1 simulation (h2) per Titan's CPU\@. 2
    simulation use 16 cores per node, 1 simulation 8. 50 Events; 1 Titan
    worker nodes; 16 work threads per node; 100 events per
    node.}\label{fig:comparison-8-16cores}
\end{figure}

A performance analysis of Titan's AMD CPUs for detector simulations also
helps to compare Titan and Grid site performance. Usually, Grid sites expose
resources with heterogeneous CPU architectures and 8 (virtual) cores, while
Titan's offer an homogeneous 16 cores architecture. We used the rate of
events processes per minute as a measure of the efficiency of executing the
same detector simulation on Titan or Grid sites. Comparisons of the
efficiencies of Titan to the BNL and SIGNET Grid sites, normalized for 8
cores, show that the effective performance per-core at Titan is
\(\approx\)0.57 event per minute, roughly 1/2 of BNL and  1/3 of SIGNET\@.
